---
title: "STA518 - The Affordability Squeeze"
subtitle: "How Rising Costs Have Outpaced Income Across American Counties (2009–2023)"
author: "Alysha Nadeem, Camden Davis, Gerrit Mitchell, Thrisha Jawahar"
date: today
format:
  html:
    self-contained: true
    embed-resources: true
    toc: true
    toc-depth: 3
    toc-title: "Table of Contents"
    theme: cosmo
execute:
  eval: true
  error: true
  echo: true
  warning: false
  message: false
---

# Overview

Between 2009 and 2023, the U.S. Consumer Price Index rose by approximately 42%, reflecting a sustained period of inflation that reshaped household budgets across the country. This report investigates whether income growth kept pace with that inflation — and whether the burden fell equally across all communities.

Using county-level American Community Survey (ACS) data merged with Bureau of Labor Statistics (BLS) CPI data, we examine trends in real median income, housing costs, and poverty rates across more than 3,000 U.S. counties. Our analysis finds that high-poverty counties experienced meaningfully lower inflation-adjusted income growth, widening the gap between the most and least economically vulnerable communities.

---

# Data Dictionary

The analysis draws on two datasets, described below:

### Census Dataset (`census_data_county_2009-2023.csv`)

*Source: U.S. Census Bureau, American Community Survey (ACS) 5-Year Estimates*

| Variable | Type | Description |
|---|---|---|
| `geoid` | Character | 5-digit Federal Information Processing Standard (FIPS) county code |
| `county_state` | Character | County and state name (e.g., "Kent County, Michigan") |
| `year` | Integer | Survey year (2009–2023) |
| `population` | Numeric | Total county population estimate |
| `median_income` | Numeric | Median household income in nominal dollars |
| `median_monthly_rent_cost` | Numeric | Median gross monthly rent cost in nominal dollars |
| `median_monthly_home_cost` | Numeric | Median selected monthly owner costs in nominal dollars |
| `prop_female` | Numeric | Proportion of county population identifying as female (0–1) |
| `prop_male` | Numeric | Proportion of county population identifying as male (0–1) |
| `prop_poverty` | Numeric | Proportion of county population below the federal poverty line (0–1) |

### CPI Dataset (`cpi_2000-2024.csv`)

*Source: Bureau of Labor Statistics, CPI-U (All Urban Consumers, All Items)*

| Variable | Type | Description |
|---|---|---|
| `series_id` | Character | BLS series identifier |
| `year` | Integer | Year of observation |
| `period` | Character | Month code (e.g., M01 for January) |
| `period_name` | Character | Month name (e.g., "January") |
| `date` | Character | Full date string |
| `cpi` | Numeric | Consumer Price Index value (base period: 1982–84 = 100) |

## Derived Variables (created during analysis)

| Variable | Type | Description |
|---|---|---|
| `geoid` | Character | 5-digit FIPS code uniquely identifying each county |
| `county_state` | Character | County and state name (e.g., "Kent County, Michigan") |
| `county_name` | Character | County name extracted from `county_state` |
| `state` | Character | State name extracted from `county_state` |
| `year` | Integer | Survey year (2009–2023) |
| `date_based_data` | Date | First day of the survey year as a Date object (parsed via lubridate) |
| `annual_avg_cpi` | Numeric | Average of 12 monthly CPI-U values for the given year (base: 1982–84 = 100) |
| `population` | Numeric | Total county population estimate at time of survey |
| `prop_female` | Numeric | Proportion of county population identifying as female (0–1) |
| `prop_male` | Numeric | Proportion of county population identifying as male (0–1) |
| `prop_poverty` | Numeric | Proportion of county population below the federal poverty line (0–1) |
| `poverty_tier` | Factor | Annual poverty quartile: Low Poverty, Medium-Low, Medium-High, High Poverty |
| `median_income` | Numeric | Median household income in nominal dollars at time of survey |
| `median_income_2024_dollars` | Numeric | Median household income adjusted to 2024 dollars using CPI ratio |
| `median_monthly_rent_cost` | Numeric | Median gross monthly rent in nominal dollars at time of survey |
| `median_monthly_rent_cost_2024_dollars` | Numeric | Median gross monthly rent adjusted to 2024 dollars |
| `median_monthly_home_cost` | Numeric | Median selected monthly owner costs in nominal dollars at time of survey |
| `median_monthly_home_cost_2024_dollars` | Numeric | Median selected monthly owner costs adjusted to 2024 dollars |
| `rent_to_income_ratio` | Numeric | Monthly rent as a share of monthly income: `median_monthly_rent / (median_income / 12)` |

---

# Data Cleaning

```{r}
#| label: load-libraries
# Core collection
library(tidyverse)

# Data wrangling / summaries
library(dplyr)
library(janitor)
library(skimr)

# Visualization
library(ggplot2)
library(ggthemes)
library(ggpubr)
library(scales)
library(viridis)
library(patchwork)

# Text analysis
library(stringr)
library(tidytext)

# Dates
library(lubridate)

# Tables / reporting
library(flextable)
library(gt)
library(knitr)

# Missing data / imputation
library(mice)
library(naniar)

# Modeling
library(tidymodels)

# Spatial
library(tigris)
library(sf)
library(leaflet)
```

```{r}
#| label: helper-functions
# Helper: view head and tail of a data frame together
ht <- function(df, n = 5) {
  cat("--- First", n, "rows ---\n")
  print(head(df, n))
  cat("\n--- Last", n, "rows ---\n")
  print(tail(df, n))
}
```

## Loading Raw Data

```{r}
#| label: load-data
# Data source URLs from course GitHub repository
census_url <- "https://raw.githubusercontent.com/dilernia/STA418-518/main/Data/census_data_county_2009-2023.csv"
cpi_url    <- "https://raw.githubusercontent.com/dilernia/STA418-518/refs/heads/main/Data/cpi_2000-2024.csv"

# Read both datasets
census_raw <- read_csv(census_url, show_col_types = FALSE)
cpi_raw    <- read_csv(cpi_url,    show_col_types = FALSE)
```

## Cleaning the Census Dataset

```{r}
#| label: clean-census
# Standardize column names using janitor
census <- census_raw |> clean_names()

# Clean and cast all variables to correct types
census_clean <- census |>
  mutate(
    # str_pad ensures all FIPS codes are exactly 5 characters (e.g., "01001")
    geoid = str_pad(as.character(geoid), width = 5, side = "left", pad = "0"),
    # str_squish removes leading/trailing/internal extra whitespace from county names
    county_state = str_squish(as.character(county_state)),
    year = as.integer(year),
    population = as.numeric(population),
    median_income = as.numeric(median_income),
    median_monthly_rent_cost = as.numeric(median_monthly_rent_cost),
    median_monthly_home_cost = as.numeric(median_monthly_home_cost),
    prop_female = as.numeric(prop_female),
    prop_male = as.numeric(prop_male),
    prop_poverty = as.numeric(prop_poverty)
  ) |>
  # Remove duplicate rows
  distinct() |>
  # Restrict to study window (2009-2023)
  filter(year >= 2009, year <= 2023)

# Remove rows with impossible values (negatives, proportions outside 0-1)
census_clean <- census_clean |>
  filter(
    is.na(population)               | population >= 0,
    is.na(median_income)            | median_income >= 0,
    is.na(median_monthly_rent_cost) | median_monthly_rent_cost >= 0,
    is.na(median_monthly_home_cost) | median_monthly_home_cost >= 0,
    is.na(prop_female)  | between(prop_female,  0, 1),
    is.na(prop_male)    | between(prop_male,    0, 1),
    is.na(prop_poverty) | between(prop_poverty, 0, 1)
  )
```

## Cleaning the CPI Dataset

```{r}
#| label: clean-cpi
# Standardize CPI column names
cpi <- cpi_raw |> clean_names()

# Clean types and remove duplicates
cpi_clean <- cpi |>
  mutate(
    year        = as.integer(year),
    period      = as.character(period),
    # str_squish trims whitespace from period_name (e.g., "January  " -> "January")
    period_name = str_squish(as.character(period_name)),
    series_id   = as.character(series_id),
    date        = as.character(date),
    cpi         = as.numeric(cpi)
  ) |>
  distinct()

# Keep only monthly records (period codes like M01, M02, ..., M12)
# str_detect used here to filter with a regex pattern
cpi_monthly <- cpi_clean |>
  filter(str_detect(period, "^M\\d{2}$"))

# Aggregate monthly CPI to yearly average to match census data structure
# ymd() from lubridate parses the date column; min() gets the first date of each year
cpi_yearly <- cpi_monthly |>
  group_by(year) |>
  summarise(
    date_based_data = min(ymd(date), na.rm = TRUE),  # lubridate: parse date string
    annual_avg_cpi  = mean(cpi, na.rm = TRUE),
    .groups = "drop"
  )
```

## Validation Checks

```{r}
#| label: validate-census
# Confirm no county appears more than once per year
census_clean |>
  count(geoid, year) |>
  filter(n > 1) |>
  nrow() |>
  cat("Duplicate county-year rows:", ... = _, "\n")

# Confirm sex proportions sum to approximately 1
census_clean |>
  mutate(sex_total = prop_female + prop_male) |>
  summarise(
    min_sex_total = min(sex_total, na.rm = TRUE),
    max_sex_total = max(sex_total, na.rm = TRUE),
    avg_sex_total = mean(sex_total, na.rm = TRUE)
  ) |>
  knitr::kable(caption = "Sex proportion totals (should be ~1.0)")
```

---

# Merging Data Sets

```{r}
#| label: merge-data
# LEFT JOIN: attach yearly CPI to every county-year row
# This preserves all census rows even if a year has no CPI match
county_fused <- census_clean |>
  left_join(cpi_yearly, by = "year")

# Verify no CPI values were lost in the join
county_fused |>
  summarise(
    total_rows       = n(),
    missing_cpi_rows = sum(is.na(annual_avg_cpi))
  ) |>
  knitr::kable(caption = "Join result: rows and CPI coverage")
```

```{r}
#| label: inflation-adjustment
# CPI for 2024 is used as the reference year for all inflation adjustments
cpi_2024 <- cpi_yearly |>
  filter(year == 2024) |>
  pull(annual_avg_cpi)

# Adjust all monetary variables to 2024 dollars using the CPI ratio
county_fused <- county_fused |>
  mutate(
    median_income_2024_dollars =
      median_income * (cpi_2024 / annual_avg_cpi),
    median_monthly_rent_cost_2024_dollars =
      median_monthly_rent_cost * (cpi_2024 / annual_avg_cpi),
    median_monthly_home_cost_2024_dollars =
      median_monthly_home_cost * (cpi_2024 / annual_avg_cpi)
  )
```

```{r}
#| label: derived-variables
# Poverty quartile tier: counties ranked annually into 4 equal groups by poverty rate
county_fused <- county_fused |>
  mutate(
    poverty_tier = ntile(prop_poverty, 4),
    poverty_tier = factor(poverty_tier,
      levels = 1:4,
      labels = c("Low Poverty", "Medium-Low", "Medium-High", "High Poverty")
    )
  )

# Extract state and county name from the combined county_state field using stringr
county_fused <- county_fused |>
  mutate(
    # str_extract pulls the part after the last comma (state name)
    state = str_extract(county_state, "[^,]+$") |> str_squish(),
    # str_extract pulls everything before the first comma (county name)
    county_name = str_extract(county_state, "^[^,]+") |> str_squish()
  )

# Compute monthly rent-to-income ratio for housing burden analysis
county_fused <- county_fused |>
  mutate(
    rent_to_income_ratio = median_monthly_rent_cost / (median_income / 12)
  )

# Reorder columns for readability
county_fused <- county_fused |>
  select(
    geoid, county_state, county_name, state, year, date_based_data,
    annual_avg_cpi, population, poverty_tier, prop_poverty,
    median_income, median_income_2024_dollars,
    median_monthly_rent_cost, median_monthly_rent_cost_2024_dollars,
    median_monthly_home_cost, median_monthly_home_cost_2024_dollars,
    rent_to_income_ratio, prop_female, prop_male
  )
```

---

# String Manipulation

The following `stringr` functions are used throughout the pipeline for meaningful data preparation:

| Function | Where Used | Purpose |
|---|---|---|
| `str_pad()` | Census cleaning | Ensure FIPS codes are exactly 5 characters with leading zeros |
| `str_squish()` | Census and CPI cleaning | Remove leading, trailing, and internal excess whitespace |
| `str_detect()` | CPI filtering; Michigan filter | Filter rows matching a regex pattern (monthly periods; Michigan counties) |
| `str_extract()` | Derived variables | Pull state and county name from a combined `county_state` string |
| `str_remove()` | Michigan chart | Strip " County" suffix from labels for cleaner axis text |

---

# Date / Time Functions

The `lubridate` package is used in two places:

1. **`ymd(date)`** — Parses the raw character date column in the CPI dataset into a proper `Date` object, enabling use of `min()` to find the first date of each year.
2. **`year(date_based_data)`** — Extracts the year integer from the parsed date for display in summary tables.

```{r}
#| label: lubridate-demo
# lubridate::year() extracts the integer year from the Date column created by ymd()
cpi_yearly |>
  mutate(display_year = year(date_based_data)) |>  # lubridate: year()
  select(display_year, annual_avg_cpi) |>
  filter(display_year >= 2009, display_year <= 2023) |>
  slice_head(n = 5) |>
  knitr::kable(caption = "Sample: year extracted from parsed date using lubridate::year()")
```

---

# Exploratory Data Analysis

## Patterns of Missingness

```{r}
#| label: missingness-summary
# Count NAs per column in the merged dataset
na_counts <- county_fused |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") |>
  filter(n_missing > 0) |>
  arrange(desc(n_missing))

na_counts |>
  knitr::kable(caption = "Variables with missing values in the merged county dataset")
```

```{r}
#| label: missingness-plot
#| fig-cap: "Missing value counts by variable. Monetary variables share a similar missingness pattern, suggesting systematic non-response in certain counties."
# Visualize missingness using naniar
gg_miss_var(county_fused,
            show_pct = TRUE) +
  labs(
    title = "Missing Data by Variable",
    subtitle = "Percentage of observations missing across all county-year rows",
    caption = "Source: U.S. Census ACS | BLS CPI-U"
  ) +
  theme_few(base_size = 13)
```

**Missingness reflection:** The primary missing variables are the monetary measures (`median_income`, `median_monthly_rent_cost`, `median_monthly_home_cost`) and their inflation-adjusted counterparts. These tend to be missing together, suggesting that some counties did not report complete ACS estimates in a given year, likely smaller, more rural counties where sample sizes are too thin for reliable estimates. This is a form of non-random missingness (MNAR): the counties most likely to be missing data are systematically different from those with complete records. As a result, any national summary statistics computed on this data will slightly overrepresent larger, urban counties, and the true picture for rural America may be worse than our charts suggest.

## Tables of Summary Statistics

```{r}
#| label: avg-nationwide-table
# National yearly averages across all counties
avg_nationwide_summaries <- county_fused |>
  group_by(date_based_data) |>
  summarise(
    AMI = mean(median_income,              na.rm = TRUE),
    MMI = mean(median_income,              na.rm = TRUE) / 12,
    MR  = mean(median_monthly_rent_cost,   na.rm = TRUE),
    MH  = mean(median_monthly_home_cost,   na.rm = TRUE),
    A_CPI = mean(annual_avg_cpi,           na.rm = TRUE),
    APP = mean(prop_poverty,               na.rm = TRUE) * 100
  )

flextable(avg_nationwide_summaries) |>
  set_header_labels(values = c(
    date_based_data = "Year",
    AMI   = "Median Income (Annual)",
    MMI   = "Median Income (Monthly)",
    MR    = "Median Rent (Monthly)",
    MH    = "Median Home Cost (Monthly)",
    A_CPI = "Annual CPI",
    APP   = "Poverty %"
  )) |>
  add_header_lines(values = "Average U.S. Monetary Information by Year") |>
  align(i = 1, align = "center", part = "header") |>
  colformat_date(j = "date_based_data", fmt_date = "%Y") |>
  colformat_double(j = c("AMI", "MMI", "MR", "MH"), prefix = "$", digits = 0, big.mark = ",") |>
  colformat_double(j = "APP",   digits = 1, suffix = "%") |>
  colformat_double(j = "A_CPI", digits = 2, big.mark = ",") |>
  autofit()
```

```{r}
#| label: cpi-summary-table
# Annual CPI distribution statistics (2009-2023)
cpi_summaries <- cpi_clean |>
  filter(year >= 2009, year < 2024) |>
  group_by(year) |>
  summarise(
    Min    = min(cpi,  na.rm = TRUE),
    Q1     = quantile(cpi, 0.25, na.rm = TRUE),
    Median = median(cpi,   na.rm = TRUE),
    Q3     = quantile(cpi, 0.75, na.rm = TRUE),
    Max    = max(cpi,  na.rm = TRUE),
    Mean   = mean(cpi, na.rm = TRUE),
    SD     = sd(cpi,   na.rm = TRUE)
  ) |>
  mutate(year = as.character(year))

flextable(cpi_summaries) |>
  set_header_labels(values = c(
    year = "Year", Min = "Min CPI", Q1 = "Q1 CPI", Median = "Median CPI",
    Q3 = "Q3 CPI", Max = "Max CPI", Mean = "Mean CPI", SD = "Std Dev"
  )) |>
  add_header_lines(values = "U.S. CPI Distribution by Year (2009–2023)") |>
  align(i = 1, align = "center", part = "header") |>
  colformat_double(j = c("Min","Q1","Median","Q3","Max","Mean"), digits = 2, big.mark = ",") |>
  colformat_double(j = "SD", digits = 3, big.mark = ",") |>
  autofit()
```

---

# Data Visualizations

## Section 1: The Inflation Context

Between 2009 and 2023, the CPI rose from approximately 214 to 305 — a 42% increase. The pace was relatively steady until 2020, when pandemic-era supply shocks accelerated inflation sharply.

```{r}
#| label: viz-cpi-trend
#| fig-cap: "U.S. CPI trend from 2000 to 2024. The shaded region marks the 2009-2023 study window. Source: Bureau of Labor Statistics."

cpi_yearly |>
  ggplot(aes(x = year, y = annual_avg_cpi)) +
  # Highlight the study window
  annotate("rect",
    xmin = 2009, xmax = 2023,
    ymin = -Inf, ymax = Inf,
    fill = "#E8F4FD", alpha = 0.6
  ) +
  geom_line(color = "grey70", linewidth = 0.8) +
  geom_point(aes(color = year >= 2009 & year <= 2023), size = 2.5, show.legend = FALSE) +
  scale_color_manual(values = c("FALSE" = "grey60", "TRUE" = "#1565C0")) +
  # Annotate the 42% rise
  annotate("segment",
    x = 2009, xend = 2023, y = 214.5, yend = 304.7,
    arrow = arrow(length = unit(0.2, "cm"), ends = "both"),
    color = "#C62828", linewidth = 0.7
  ) +
  annotate("label",
    x = 2016, y = 245,
    label = "+42% (2009-2023)",
    color = "#C62828", size = 4, fontface = "bold",
    fill = "white", label.size = 0.3
  ) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 2)) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0.02, 0.05))) +
  labs(
    title = "U.S. Consumer Price Index, 2000-2024",
    subtitle = "Shaded region (2009-2023) corresponds to the census data window",
    x = "Year",
    y = "Annual Average CPI",
    caption = "Source: Bureau of Labor Statistics, CPI-U"
  ) +
  theme_few(base_size = 14) +
  theme(panel.grid.minor = element_blank())
```

## Section 2: Did Incomes Keep Up?

High-poverty counties started lower and grew more slowly. By 2023, the gap in real median income between Low Poverty and High Poverty counties had widened, meaning the affordability squeeze landed hardest on the communities least able to absorb it. The post-2020 acceleration in income for Low Poverty counties reflects labor market tightening after the pandemic, an effect that did not reach poorer communities equally.

```{r}
#| label: viz-income-trend
#| fig-cap: "Inflation-adjusted median income by county poverty tier, 2009-2023. All values in 2024 dollars. Source: U.S. Census ACS, BLS CPI-U."

# Aggregate to national yearly averages by poverty tier
income_trend <- county_fused |>
  filter(!is.na(median_income_2024_dollars), !is.na(poverty_tier)) |>
  group_by(year, poverty_tier) |>
  summarise(
    avg_real_income = mean(median_income_2024_dollars, na.rm = TRUE),
    n_counties = n(),
    .groups = "drop"
  )

# Color-blind-friendly palette: blue for low poverty, orange/red for high
tier_colors <- c(
  "Low Poverty"  = "#1565C0",
  "Medium-Low"   = "#42A5F5",
  "Medium-High"  = "#EF6C00",
  "High Poverty" = "#B71C1C"
)

income_trend |>
  ggplot(aes(x = year, y = avg_real_income, color = poverty_tier, group = poverty_tier)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  # Direct labels at 2023 endpoints
  geom_text(
    data = income_trend |> filter(year == 2023),
    aes(label = poverty_tier),
    hjust = -0.08, size = 3.2, fontface = "bold"
  ) +
  scale_color_manual(values = tier_colors) +
  scale_x_continuous(
    breaks = c(2009, 2012, 2015, 2017, 2019, 2021, 2023),
    expand = expansion(mult = c(0.02, 0.22))
  ) +
  scale_y_continuous(labels = scales::dollar_format(scale = 1e-3, suffix = "K")) +
  labs(
    title = "Inflation-Adjusted Median Income by County Poverty Tier (2009-2023)",
    subtitle = "All values in 2024 dollars. Poverty tiers assigned annually by quartile.",
    x = "Year",
    y = "Average Real Median Income (2024 $)",
    caption = "Source: U.S. Census ACS | BLS CPI-U"
  ) +
  theme_few(base_size = 14) +
  theme(legend.position = "none")
```

## Section 3: The Housing Burden

The 30% rule of thumb holds that households spending more than 30% of their gross income on rent are housing-cost-burdened. The scatter below shows that in 2023, many low-income (high-poverty) counties fall above or near this threshold. Michigan's highest-cost counties skew toward wealthier, lower-poverty areas. The burden there is driven by rising market prices rather than income deprivation.

```{r}
#| label: viz-housing-burden
#| fig-cap: "Left: Rent vs. income across all U.S. counties in 2023, colored by poverty rate. Points above the dashed line face housing cost burden. Right: Top 10 Michigan counties by real monthly home cost. Source: U.S. Census ACS | BLS CPI-U."
#| fig-width: 14
#| fig-height: 6

# National scatter: rent burden vs income, 2023
scatter_2023 <- county_fused |>
  filter(
    year == 2023,
    !is.na(median_monthly_rent_cost_2024_dollars),
    !is.na(median_income_2024_dollars),
    !is.na(prop_poverty)
  )

p_scatter <- scatter_2023 |>
  ggplot(aes(
    x = median_income_2024_dollars,
    y = median_monthly_rent_cost_2024_dollars,
    color = prop_poverty
  )) +
  geom_point(alpha = 0.5, size = 1.4) +
  # 30% burden threshold line
  geom_abline(
    slope = 0.30 / 12, intercept = 0,
    linetype = "dashed", color = "#C62828", linewidth = 0.9
  ) +
  annotate("text",
    x = 155000, y = 3550,
    label = "30% burden\nthreshold",
    color = "#C62828", size = 3.2, hjust = 1
  ) +
  scale_color_viridis_c(option = "C", labels = scales::percent, name = "Poverty Rate") +
  scale_x_continuous(labels = scales::dollar_format(scale = 1e-3, suffix = "K")) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    title = "Rent Burden vs. Median Income, U.S. Counties (2023)",
    subtitle = "Points above dashed line are housing-cost-burdened.",
    x = "Median Annual Income (2024 $)",
    y = "Median Monthly Rent (2024 $)",
    caption = "Source: U.S. Census ACS | BLS CPI-U"
  ) +
  theme_few(base_size = 12)

# Michigan regional callout: top 10 counties by real home cost
michigan_2023 <- county_fused |>
  filter(
    str_detect(county_state, regex("michigan", ignore_case = TRUE)),
    year == 2023,
    !is.na(median_monthly_home_cost_2024_dollars)
  ) |>
  # str_remove strips " County" from label for cleaner axis text
  mutate(county_label = str_remove(county_name, " County")) |>
  arrange(desc(median_monthly_home_cost_2024_dollars)) |>
  slice_head(n = 10)

p_michigan <- michigan_2023 |>
  ggplot(aes(
    x = median_monthly_home_cost_2024_dollars,
    y = fct_reorder(county_label, median_monthly_home_cost_2024_dollars),
    fill = prop_poverty
  )) +
  geom_col() +
  scale_fill_viridis_c(option = "C", labels = scales::percent, name = "Poverty Rate") +
  scale_x_continuous(
    labels = scales::dollar_format(),
    expand = expansion(mult = c(0, 0.1))
  ) +
  labs(
    title = "Top 10 Michigan Counties: Real Monthly Home Costs (2023)",
    subtitle = "Inflation-adjusted to 2024 dollars.",
    x = "Median Monthly Home Cost (2024 $)",
    y = NULL,
    caption = "Source: U.S. Census ACS | BLS CPI-U"
  ) +
  theme_few(base_size = 12)

# Combine side by side using patchwork
p_scatter + p_michigan + plot_layout(widths = c(1.6, 1))
```

## Section 4: Where It Hurts Most

The two maps below show the geographic concentration of poverty and rent burden at the county level for 2023. High poverty is concentrated in the Deep South, Appalachian corridor, and parts of the Southwest. The rent burden map reveals a different geography — coastal metros and mountain resort counties carry the highest ratios, reflecting markets where rent outpaces income regardless of poverty rate.

### Map A: County Poverty Rates (Interactive)

```{r}
#| label: map-setup
#| cache: true
#| include: false
options(tigris_use_cache = TRUE)

# Load county shapefiles and join to 2023 data
county_shape <- tigris::counties(cb = TRUE, year = 2023, class = "sf")

map_data_2023 <- county_fused |> filter(year == 2023)

gg_map_2023 <- county_shape |>
  left_join(map_data_2023, by = c("GEOID" = "geoid")) |>
  filter(STUSPS %in% state.abb) |>
  tigris::shift_geometry() |>
  st_cast("MULTIPOLYGON") |>
  st_transform(crs = 4326)

# Simplify geometry for performance
gg_map_2023_simple <- gg_map_2023 |>
  st_simplify(dTolerance = 0.01, preserveTopology = TRUE)

# Color palettes
pal_poverty <- colorNumeric(
  palette  = "YlOrRd",
  domain   = gg_map_2023_simple$prop_poverty,
  na.color = "white"
)
```

```{r}
#| label: leaflet-poverty
#| fig-height: 5
#| out-width: "100%"

leaflet(gg_map_2023_simple, options = leafletOptions(preferCanvas = TRUE)) |>
  addProviderTiles(providers$CartoDB.PositronNoLabels) |>
  addPolygons(
    fillColor   = ~pal_poverty(prop_poverty),
    fillOpacity = 0.8,
    color       = "grey80",
    weight      = 0.4,
    smoothFactor = 0.5,
    highlightOptions = highlightOptions(
      weight = 1.5, color = "#333333", fillOpacity = 0.95, bringToFront = TRUE
    ),
    label = ~lapply(paste0(
      "<b>", county_state, "</b><br>",
      "Poverty Rate: ", scales::percent(prop_poverty, accuracy = 0.1), "<br>",
      "Median Income: ", scales::dollar(median_income, accuracy = 1)
    ), htmltools::HTML)
  ) |>
  addLegend(
    position  = "bottomright", pal = pal_poverty,
    values    = ~prop_poverty, title = "Poverty Rate",
    labFormat = labelFormat(suffix = "%", transform = function(x) round(x * 100, 1))
  ) |>
  addControl(
    html     = "<b>County Poverty Rates (2023)</b><br><small>Source: U.S. Census ACS</small>",
    position = "topleft"
  )
```

### Map B: Rent-to-Income Ratio (Static, County Level)

```{r}
#| label: viz-rent-burden-map
#| fig-cap: "County-level rent-to-income ratio in 2023 (static view). Values above 0.30 indicate rent burden; capped at 0.60 for color scaling. Alaska and Hawaii shifted for display. Source: U.S. Census ACS."
#| fig-width: 11
#| fig-height: 6

# Static ggplot choropleth for rent burden (avoids second interactive map)
gg_map_2023 |>
  mutate(rent_burden_capped = pmin(rent_to_income_ratio, 0.60, na.rm = TRUE)) |>
  ggplot(aes(fill = rent_burden_capped)) +
  geom_sf(color = "#C0C0C0", linewidth = 0.1) +
  scale_fill_gradient2(
    low      = "#E8F5E9",
    mid      = "#FFF176",
    high     = "#B71C1C",
    midpoint = 0.30,
    limits   = c(0, 0.60),
    labels   = scales::percent,
    name     = "Rent / Income\n(capped at 60%)",
    na.value = "grey90"
  ) +
  labs(
    title    = "Monthly Rent as a Share of Income, U.S. Counties (2023)",
    subtitle = "Red counties spend 30%+ of monthly income on rent. Values capped at 60% for color scale.",
    caption  = "Source: U.S. Census ACS | BLS CPI-U"
  ) +
  theme_void(base_size = 13) +
  theme(legend.position = "right")
```

## Section 5: Income Growth Distribution by Poverty Tier

Before testing whether differences are statistically significant, it is useful to see the raw distributions of income growth. High-poverty counties are clustered near zero or negative growth, while low-poverty counties show a wider right tail.

```{r}
#| label: viz-income-growth-dist
#| fig-cap: "Distribution of real income growth (2009-2023) by poverty tier. High-poverty counties are concentrated near zero or negative growth. Source: U.S. Census ACS | BLS CPI-U."

income_growth <- county_fused |>
  filter(year %in% c(2009, 2023), !is.na(median_income_2024_dollars)) |>
  select(geoid, year, median_income_2024_dollars, poverty_tier) |>
  pivot_wider(
    names_from  = year,
    values_from = median_income_2024_dollars,
    names_prefix = "income_"
  ) |>
  filter(!is.na(income_2009), !is.na(income_2023)) |>
  mutate(income_growth_dollars = income_2023 - income_2009)

income_growth |>
  filter(!is.na(poverty_tier)) |>
  ggplot(aes(x = income_growth_dollars, fill = poverty_tier)) +
  geom_histogram(bins = 50, color = "white", alpha = 0.85) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black", linewidth = 0.7) +
  facet_wrap(~poverty_tier, ncol = 2, scales = "free_y") +
  scale_fill_manual(values = c(
    "Low Poverty"  = "#1565C0",
    "Medium-Low"   = "#42A5F5",
    "Medium-High"  = "#EF6C00",
    "High Poverty" = "#B71C1C"
  )) +
  scale_x_continuous(labels = scales::dollar_format(scale = 1e-3, suffix = "K")) +
  labs(
    title    = "Distribution of Real Income Growth by County Poverty Tier (2009-2023)",
    subtitle = "Dashed line at $0: counties to the left saw negative real income growth",
    x        = "Change in Real Median Income (2024 $)",
    y        = "Number of Counties",
    caption  = "Source: U.S. Census ACS | BLS CPI-U"
  ) +
  theme_few(base_size = 13) +
  theme(legend.position = "none")
```

---

# Permutation Test

## Hypotheses

We test whether counties in the **highest poverty quartile** experienced meaningfully different inflation-adjusted income growth from 2009 to 2023 compared to counties in the **lowest poverty quartile**.

- **Null hypothesis (H0):** The mean real income growth from 2009 to 2023 is the same for High Poverty and Low Poverty counties.
- **Alternative hypothesis (H1):** The mean real income growth differs between High Poverty and Low Poverty counties.

The test statistic is the difference in group means: `mean(High Poverty growth) - mean(Low Poverty growth)`.

## Permutation Procedure

```{r}
#| label: permutation-test

# Keeping only bottom and top poverty quartile counties
growth_filtered <- income_growth |>
  filter(poverty_tier %in% c("Low Poverty", "High Poverty"))

# Observed difference in mean real income growth
obs_diff <- growth_filtered |>
  group_by(poverty_tier) |>
  summarise(mean_growth = mean(income_growth_dollars, na.rm = TRUE), .groups = "drop") |>
  summarise(diff = diff(mean_growth)) |>
  pull(diff)

cat("Observed mean growth difference (High minus Low Poverty):", scales::dollar(round(obs_diff, 0)), "\n")

# Permutation test: shuffle poverty tier labels 5,000 times to build null distribution
set.seed(518)
n_perms <- 5000

perm_diffs <- replicate(n_perms, {
  shuffled <- growth_filtered |>
    mutate(poverty_tier = sample(poverty_tier))

  shuffled |>
    group_by(poverty_tier) |>
    summarise(mean_growth = mean(income_growth_dollars, na.rm = TRUE), .groups = "drop") |>
    summarise(diff = diff(mean_growth)) |>
    pull(diff)
})

# Two-sided p-value
p_value <- mean(abs(perm_diffs) >= abs(obs_diff))
cat("Permutation test p-value:", round(p_value, 4), "\n")

# Alpha cutoff for visual reference (two-sided, alpha = 0.05)
alpha_cutoff <- quantile(abs(perm_diffs), 0.95)
```

## Null Distribution

```{r}
#| label: viz-permutation
#| fig-cap: "Null distribution from 5,000 permutations. The red dashed line marks the observed difference; the blue dotted lines mark the alpha = 0.05 cutoff. Source: U.S. Census ACS | BLS CPI-U."

tibble(perm_diff = perm_diffs) |>
  ggplot(aes(x = perm_diff)) +
  geom_histogram(bins = 60, fill = "#90CAF9", color = "white") +
  # Observed test statistic
  geom_vline(
    xintercept = obs_diff,
    color = "#C62828", linewidth = 1.2, linetype = "dashed"
  ) +
  # Alpha = 0.05 cutoff lines (two-sided)
  geom_vline(
    xintercept = c(-alpha_cutoff, alpha_cutoff),
    color = "#1565C0", linewidth = 0.9, linetype = "dotted"
  ) +
  annotate("label",
    x = obs_diff, y = Inf,
    label = paste0("Observed\n", scales::dollar(round(obs_diff, 0))),
    vjust = 1.4, color = "#C62828", size = 3.5,
    fill = "white", label.size = 0.3
  ) +
  annotate("text",
    x = min(perm_diffs) * 0.85, y = Inf,
    label = paste0("p = ", round(p_value, 4)),
    vjust = 1.8, hjust = 0, size = 4, fontface = "bold"
  ) +
  annotate("text",
    x = alpha_cutoff * 1.02, y = Inf,
    label = "alpha = 0.05\ncutoff",
    vjust = 1.8, hjust = 0, color = "#1565C0", size = 3.2
  ) +
  scale_x_continuous(labels = scales::dollar_format()) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.08))) +
  labs(
    title    = "Permutation Test: Income Growth Difference by Poverty Tier (2009-2023)",
    subtitle = "Null distribution from 5,000 random shuffles of poverty tier labels",
    x        = "Difference in Mean Real Income Growth (High minus Low Poverty)",
    y        = "Count",
    caption  = "Source: U.S. Census ACS | BLS CPI-U"
  ) +
  theme_few(base_size = 14)
```

## Interpretation

The observed mean income growth gap between High Poverty and Low Poverty counties was `r scales::dollar(round(obs_diff, 0))`, with a permutation p-value of `r round(p_value, 4)`. Because this p-value is less than 0.05, we reject the null hypothesis. The observed difference falls well outside the range of differences that would be expected by chance alone (red dashed line is far beyond the blue dotted alpha cutoff lines). This provides strong statistical evidence that High Poverty counties experienced meaningfully lower real income growth over the study period. Therefore, the affordability squeeze was not distributed equally.

---

# Conclusions / Main Takeaways

1. **Prices rose 42% from 2009 to 2023.** The CPI trend confirms a sustained inflationary environment, accelerating after 2020.

2. **Incomes did not keep pace equally.** Inflation-adjusted median income grew in all poverty tiers, but the gap between Low and High Poverty counties widened over the period. High-poverty counties ended 2023 further behind in real terms than they started.

3. **Housing cost burden is concentrated in high-poverty counties.** The national scatter (Section 3) shows that counties above the 30% rent burden threshold cluster among those with higher poverty rates and lower incomes.

4. **Geography matters.** Poverty is concentrated in the South and Appalachia; rent burden is a coastal and resort-market phenomenon. These are distinct housing crises requiring distinct policy responses.

5. **The income growth gap is statistically significant.** The permutation test confirms (p < 0.05) that the lower income growth observed in High Poverty counties is not attributable to chance. The affordability squeeze was systematically worse for the communities least equipped to absorb it.

---

# Contributions of Each Group Member

| Member | Contributions |
|---|---|
| Alysha Nadeem | Report structure and narrative; data visualizations; missingness analysis; permutation test; conclusions |
| Thrisha Jawahar | Data cleaning and preprocessing; dataset merging; inflation adjustment calculations |
| Gerrit Mitchell | Data cleaning and preprocessing; diagnostic summaries; summary statistics tables; data dictionaries |
| Camden Davis | Collaboration infrastructure; CPI and Michigan visualizations |

---

# Appendix: Diagnostic and Validation Code

The following code blocks were used during the analysis phase for data validation and exploration.

## Census Raw Data Overview

```{r}
#| label: appendix-census-glimpse
glimpse(census_raw)
```

```{r}
#| label: appendix-census-skim
skim(census_raw)
```

```{r}
#| label: appendix-census-ht
ht(census_raw)
```

## CPI Raw Data Overview

```{r}
#| label: appendix-cpi-glimpse
glimpse(cpi_raw)
```

```{r}
#| label: appendix-cpi-ht
ht(cpi_raw)
```

## Cleaned Data Overviews

```{r}
#| label: appendix-clean-glimpse
glimpse(census_clean)
glimpse(cpi_clean)
```

## CPI Monthly Records Check

```{r}
#| label: appendix-cpi-monthly-check
# Verify 12 months per year in filtered monthly data
cpi_monthly |>
  count(year) |>
  print(n = 30)
```

## Merged Dataset Overview

```{r}
#| label: appendix-fused-overview
glimpse(county_fused)
skim(county_fused)
```

## Row Count by Year

```{r}
#| label: appendix-rows-per-year
county_fused |>
  count(year) |>
  print(n = 20)
```

## Impossible Value Checks

```{r}
#| label: appendix-impossible-values
census_clean |>
  summarise(
    negative_population      = sum(population < 0,              na.rm = TRUE),
    negative_income          = sum(median_income < 0,           na.rm = TRUE),
    negative_rent            = sum(median_monthly_rent_cost < 0, na.rm = TRUE),
    negative_home            = sum(median_monthly_home_cost < 0, na.rm = TRUE),
    female_out_of_bounds     = sum(prop_female < 0 | prop_female > 1, na.rm = TRUE),
    male_out_of_bounds       = sum(prop_male   < 0 | prop_male   > 1, na.rm = TRUE),
    poverty_out_of_bounds    = sum(prop_poverty < 0 | prop_poverty > 1, na.rm = TRUE)
  ) |>
  knitr::kable(caption = "Count of impossible values after cleaning (should all be 0)")
```
